---
title: "PS270 Final Project"
author: "Cecilia Lees"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---

## Project thoughts

I am interested in exploring data related to gender and politics in the United States. Some ideas I had were comparing the amount of women in different levels or in different branches of government, or analyzing public opinion among various groups about women in political leadership positions. These ideas are very broad and would likely include large sets of data, but I hope to narrow them down into a more specific topic that relates to women in politics as I continue to explore what data is out there. I also would be interest in exploring topic related to the environment.

Research question: How does a country's level of democracy affect its Environmental Performance Index (EPI)?

My hypothesis is that countries with higher levels of democracy have a higher EPI. This is because countries with higher levels of democracy have higher levels of public participation, which I think will allow for more attention to be brought to environmental issues. This will then lead to democracies being more likely to enact policies and programs in support of the environment.

My explanatory variable is the democracy rating of each country. The data I use for this variable is produced by the Economist Intelligence Unit (EIU) which ranks countries on a scale from 0 to 10, with 10 being the most democratic. The EIU explains that the democracy index is determined by factors such as the "Extent to which citizens can choose their political leaders in free and fair elections, enjoy civil liberties, prefer democracy over other political systems, can and do participate in politics, and have a functioning government that acts on their behalf".

For my outcome variable of interest I am looking at the EPI of each country. The EPI ranks countries on a scale from 0-100, with 0 being the worst performance and 100 being the best performance. The EPI ranks countries based on "climate change performance, environmental health, and ecosystem vitality". 

In order to support my hypothesis, the data would need to show that countries with higher scores of democracy have higher EPI scores.


```{r}
library(readr)
library(tidyverse)
EPI_data <- read_csv("https://epi.yale.edu/downloads/epi2024results.csv")
```

```{r}
glimpse(EPI_data)
```

```{r}
EPI_data <- EPI_data |> select(iso, country, EPI.new)
EPI_data
```

```{r}
library(tidyverse)
democracy_data <- read_csv("~/Downloads/democracy-index-eiu/democracy-index-eiu.csv")
glimpse(democracy_data)
```

```{r}
democracy_data <- democracy_data |> filter(Year==2024)
democracy_data
```

```{r}
library(tidyverse)
democracy_data <- arrange(democracy_data, `Democracy Index`)
democracy_data
```

```{r}
democracy_data <- democracy_data |> rename(country = Entity)
democracy_data
```

```{r}
combined_data <- inner_join(democracy_data,EPI_data,by = "country")
combined_data
```

```{r}
graph_data <- combined_data |> group_by(`World region according to OWID`) |> summarize(avg_dem_index = mean(`Democracy Index`), avg_EPI_index = mean(EPI.new))
graph_data
```

```{r}
graph_data |> ggplot(aes(x = `World region according to OWID`, y = avg_dem_index)) + geom_col() + labs(x = "Continent", y = "Average Democracy Score", title = "Average Democracy Score by World Region")

graph_data |> ggplot(aes(x = `World region according to OWID`, y = avg_EPI_index)) + geom_col() + labs(x = "Continent", y = "Average EPI", title = "Average EPI by World Region")
```